Open archaeology, open source?

### 2024-03-29 analysis/_02-background.qmd
More updates in response to reviewer comments.

Open science is typically compared with the open source movement in that they both involve a distributed, digitally-mediated and worldwide labour force, who somehow derive rough consensus directed towards assets held in the public domain [@tennant2020]. But they differ in terms of the contexts in which they operate, the stakeholders involved, and the kinds of outcomes they produce. Whereas open source emerged from concerns over consumer rights and then developed as a means of maintaining resilient and collectively motivated projects, open science comes out of a desire to make research practices more transparent and accessible. Open source is performed by professional and hobbyist software developers alike, and participants contribute in a wide variety of ways (including: programming, writing documentation, translating software and documentation, bug reporting, and financial support), but, in open science, scientists are usually the only participants actively involved in creating and maintaining contributions. 
Moreover, whereas open source projects often attract participants with varied stakes in the software and use cases in mind, open science projects are typically bounded by small communities of specialists with very particular needs [@kling2003]. Additionally, open science is bounded by the professional contexts in which science operates, and as such, produces outputs that can be easily credited to specific sets of individuals for reasons of resume-building, tenure and promotion [@mirowski2018; @dorta-gonzalez2021]. Open science projects whose contributions are supported by research funding also face sustainability concerns, as participants lose motivation to contribute once funding runs out [@carver2022; @adema2021]. Once a project is completed, papers have been published, and credit has been allocated, it is common for scientists to mark their projects as finished and move on to new endeavours [@kelty2008: 271-275; @howison2013]. Scientists may archive their work in an institutional repository, at which point the work enters a stasis state that invites reference to the completed work, but which precludes any potential for the work to be updated, modified, or directly built upon. Open source projects, on the other hand, are motivated by a more practical need for the software to function properly in perpetuity, and contributors may remain actively or sporadically involved to satisfy users' needs, or to direct users to derivative and functional forks of abandoned software [@kelty2008: 278-281; @coleman2012: 116-122; @hippel2003].

@@

Software forges—collaborative web platforms like GitHub, GitLab and Codeberg—are designed to facilitate open source software development by hosting public git repositories.
However, they also support common software developer and project management practices, such as issue and bug tracking, code-commenting, task management, identity and permissions management, web publishing, vulnerability detection, creation and maintenance of metadata, and financial sponsorship.^[Much of this relies on proprietary software owned and managed by commercial organisations, and there has been some controversy about the take-over of open source platforms by for-profit entities (see XXXX)].
These platforms also implement standard social media functions, like the ability to follow projects and individual users to receive updates on their activities, "star" certain repositories as a combine bookmarking and 'like' feature, and maintain a public-facing profile that includes personally-identifying information (e.g. profile picture, username, real name, employer or affiliation), references to all public activity on the platform, and links to the user's other social media profiles.
Code-sharing platforms thus serve as comprehensive developer portfolios and community networking resources.
While these additional features are meant to complement and enhance the experience of contributing to open source projects, they are not actually part of the git protocol.

>>>>

Open science is typically compared with the open source movement in that they both involve a distributed, digitally-mediated and worldwide labour force, who somehow derive rough consensus directed towards assets held in the public domain [@tennant2020]. But they differ in terms of the contexts in which they operate, the stakeholders involved, and the kinds of outcomes they produce. Whereas open source emerged from concerns over consumer rights and then developed as a means of maintaining resilient and collectively motivated projects, open science comes out of a desire to make research practices more transparent and accessible. Open source is performed by professional and hobbyist software developers alike, and participants contribute in a wide variety of ways (including: programming, writing documentation, translating software and documentation, bug reporting, and financial support), but, in open science, scientists are usually the only participants actively involved in creating and maintaining contributions.^[Though, participatory or community-based research does exist and is an exception to this generalization (see @morgan2012). It requires active effort to do well and should not be considered a passive by-product of posting one's research on the web.]
Moreover, whereas open source projects often attract participants with varied stakes in the software and use cases in mind, open science projects are typically bounded by small communities of specialists with very particular needs [@kling2003]. Additionally, open science is bounded by the professional contexts in which science operates, and as such, produces outputs that can be easily credited to specific sets of individuals for reasons of resume-building, tenure and promotion [@mirowski2018; @dorta-gonzalez2021]. Open science projects whose contributions are supported by research funding also face sustainability concerns, as participants lose motivation to contribute once funding runs out [@carver2022; @adema2021]. Once a project is completed, papers have been published, and credit has been allocated, it is common for scientists to mark their projects as finished and move on to new endeavours [@kelty2008: 271-275; @howison2013]. Scientists may archive their work in an institutional repository, at which point the work enters a stasis state that invites reference to the completed work, but which precludes any potential for the work to be updated, modified, or directly built upon. Open source projects, on the other hand, are motivated by a more practical need for the software to function properly in perpetuity, and contributors may remain actively or sporadically involved to satisfy users' needs, or to direct users to derivative and functional forks of abandoned software [@kelty2008: 278-281; @coleman2012: 116-122; @hippel2003].

@@

Software forges—collaborative web platforms like GitHub, GitLab and Codeberg—are designed to facilitate open source software development by hosting public git repositories.
However, they also support common software developer and project management practices, such as issue and bug tracking, code-commenting, task management, identity and permissions management, web publishing, vulnerability detection, creation and maintenance of metadata, and financial sponsorship.^[Much of this relies on proprietary software owned and managed by commercial organisations, and there has been some controversy surrounding the take-over of open source platforms by for-profit entities (see @saunders2022 and @brembs2023)].
These platforms also implement standard social media functions, like the ability to follow projects and individual users to receive updates on their activities, "star" certain repositories as a combine bookmarking and 'like' feature, and maintain a public-facing profile that includes personally-identifying information (e.g. profile picture, username, real name, employer or affiliation), references to all public activity on the platform, and links to the user's other social media profiles.
Code-sharing platforms thus serve as comprehensive developer portfolios and community networking resources.
While these additional features are meant to complement and enhance the experience of contributing to open source projects, they are not actually part of the git protocol.

### 2024-03-29 analysis/_03-data_methodology.qmd
More updates in response to reviewer comments.

We opted to only collect repository data from GitHub because it is the most popular forge platform used by open-archaeo projects (@tbl-forges).
This means that projects that do not use version control (`r percent(p_no_vc)` of the total), or which develop and host code on platforms other than GitHub (`r percent(p_no_github)` of the total), were excluded from parts of the analysis that examine *how* archaelogists develop software.
However, we were still able to draw from all records to ascertain the general composition of open-archaeo, and by extension, to address *what* kinds of software and resources archaeologists make.

>>>>

We opted to only collect repository data from GitHub because it is the most popular forge platform used by open-archaeo projects (@tbl-forges).
This means that we excluded projects that do not use version control (`r percent(p_no_vc)` of the total) or which develop and host code on platforms other than GitHub (`r percent(p_no_github)` of the total) from the parts of the analysis that examine *how* archaelogists develop software.
However, we were still able to draw from all records to ascertain the general composition of open-archaeo, and by extension, to address *what* kinds of software and resources archaeologists make.

### 2024-03-29 analysis/_05-collaboration.qmd
Comment on new conclusion in section 5
More updates in response to reviewer comments.

The prototypical open source project comprises a core group of developers (often a single maintainer) that regularly commit new code, a wider network of collaborators that contribute through forks and pull requests, plus an active user base that create and comment on issues, who have indicated their support for the project by starring its repository.
It is unclear whether archaeological software developers actually aim to operate following this model, or whether it is even suitable for supporting what open science aims to achieve.
However it is clear that only a small number of open-archaeo projects operate according to this model.
The majority of projects are in fact short-lived, with few contributors and a small number of commits.
Use of GitHub's collaboration features is also generally low (@fig-collab-features), although the data also shows a divergence between the uptake of features that facilitate direct code contributions (forks, issues, pull requests), which have markedly zero-skewed distributions, versus more indirect, social media-like features (comments, stars), which are moderately well-used.
We hypothesise that this shows a preference for passive/reactive rather than active/proactive engagement with others work – a point we will return to in the conclusion.

>>>>

The prototypical open source project comprises a core group of developers (often a single maintainer) that regularly commit new code, a wider network of collaborators that contribute through forks and pull requests, plus an active user base that create and comment on issues, who have indicated their support for the project by starring its repository.
It is unclear whether archaeological software developers actually aim to operate following this model, or whether it is even suitable for supporting what open science aims to achieve.
However it is clear that only a small number of open-archaeo projects operate according to this model.
The majority of projects are in fact short-lived, with few contributors and a small number of commits.
Use of GitHub's collaboration features is also generally low (@fig-collab-features), although the data also shows a divergence between the uptake of features that facilitate direct code contributions (forks, issues, pull requests), which have markedly zero-skewed distributions, versus more indirect, social media-like features (comments, stars), which are moderately well-used.

These findings show a preference for passive/reactive rather than active/proactive engagement with others' work, which is not conducive to achieving the desired outcomes of opening source code, namely enabling greater engagement, reuse and critique.
While limited time or technical capability may be contributing factors (which should be targeted in more focused investigations), we believe that social norms, such as expectations and taboos surrounding the permeability of a grant-funded project's boundaries, or whether it is proper to actively engage or interfere with the work directed under another the aegis of another project, play a very significant role.
Our network analysis of collaborative ties, presented in the following section, corroborates this claim.

### 2024-03-29 analysis/_07-conclusion.qmd
More updates in response to reviewer comments.

As such, we advocate for more focused attention on specific disciplinary norms and institutional support structures that inform how knowledge is created and validated, and how varied contributions to the scholarly enterprise are mediated, credited, and valued. 
These factors, which are often ignored in open science discourse, vary from discipline to discipline and are rarely accounted for in the infrastructures themselves and in the policies that dictate or guide their implementation. 
However these factors should be considered as equal, if not more important, elements contributing to the success of open science, relative to the technical apparatus.

>>>>

As such, we advocate for more focused attention on specific disciplinary norms and institutional support structures that inform how knowledge is created and validated, and how varied contributions to the scholarly enterprise are mediated, credited, and valued.
In other words, if we want to make open source effective in relation to the aforementioned goals of encouraging greater inclusivity, transparency, and productivity, we also need to foster a culture that supports active, pragmatic and humble critique, and which instills a de-territorialized attitude concerning what it means to contribute to collective knowledge [@morgan2015; @ducke2015].
This means fighting against the pathological power-relations that scaffold all aspects of academic life, and not fooling ourselves into believing that technical solutions (i.e. using git) will, on their own, resolve the wicked social problems that lie at the heart of scientific research practice.
